---
title: "Interacción entre niveles"
date: "2023-10-13"
lang: es
number-sections: true
website:
  comments:
    giscus:
      repo: cursos-metodos-facso/multinivel
---

```{r setup, include=FALSE, cache = FALSE}
require("knitr")
opts_chunk$set(warning=FALSE,
             message=FALSE,
             echo=TRUE,
             cache = FALSE, fig.width=7, fig.height=5.2)
knitr::opts_knit$set(root.dir = here::here())
```


# Cargar/instalar librerías
```{r warning=FALSE}
pacman::p_load(lme4,haven,foreign, stargazer, texreg, lattice, sjPlot, dplyr, ggplot2, ggeffects) # paquetes a cargar
```

# Ejemplo 1, datos HSB

## Datos

```{r results='hide', message=FALSE}
mlm = read_dta("http://www.stata-press.com/data/mlmus3/hsb.dta")
dim(mlm)
names(mlm)
attach(mlm)
```

Tabla estadisticos descriptivos
```{r}
stargazer(as.data.frame(mlm),title="Estadísticos descriptivos", type = "text")
```

## Modelos

- Revisar estructura variable para la interacción

```{r results='asis'}

str(mlm$sector)
table(mlm$sector)
mlm$sector_f=as.factor(mlm$sector) # Cambiar a factor
```

- Pendiente fija & aleatoria

```{r}
reg_mlm3a = lmer(mathach ~ 1 + ses + sector_f +  mnses + (1 | schoolid),data=mlm)

# Con pendiente aleatoria
reg_mlm3b = lmer(mathach ~ 1 + ses + sector_f +  mnses + (1 + ses | schoolid),data=mlm)
```


- Devianza (ajuste comparativo pendiente fija vs aleatoria)

```{r}
anova(reg_mlm3b,reg_mlm3a)
```


- Estimar modelo con interacción

```{r}
reg_mlm3c = lmer(mathach ~ 1 + ses + sector_f + ses*sector_f + mnses +  (1 + ses | schoolid), data=mlm)
screenreg(c(reg_mlm3a,reg_mlm3b, reg_mlm3c))

```

## Plot

```{r}
plot_model(reg_mlm3c, type = "int")
```


# Ejemplo Aguinis

- Link a paper [aquí](https://multinivel.netlify.com/readings/aguinis2013crosslevel.pdf)

- Planteamiento general:

" Overall, Chen et al.’s theoretical model
predicted that employees who report higher LMX (i.e., a better relationship with their leader)
will feel more empowered (i.e., they have the autonomy and capability to perform meaning-
ful work that can affect their organization). In addition, Chen et al.’s model included the
hypothesis that the team-level variable leadership climate (i.e., ambient leadership behaviors
directed at the team as a whole) would also affect individual-level empowerment positively.
Moreover, Chen et al. hypothesized that the relationship between LMX and empowerment
would be moderated by leadership climate such that the relationship would be stronger for
teams with a better leadership climate." (p.1492)

- Preguntas e hipótesis:

1.	 Lower-level direct effects. Does a lower-level predictor X (i.e., Level 1 or L1 predictor) have an
effect on a lower-level outcome variable Y (i.e., L1 outcome)? Specifically regarding our illus-
tration, there is an interest in testing whether LMX, as perceived by subordinates, predicts
individual empowerment. Note that LMX scores are collected for each individual worker (i.e.,
there is no aggregation of such scores for the purpose of testing the presence of a lower-level
direct effect).

2.	 Cross-level direct effects. Does a higher-level predictor W (i.e., Level 2 or L2 predictor) have an
effect on an L1 outcome variable Y? Specifically, we would like to assess whether L2 variable
leadership climate predicts L1 outcome individual empowerment.

3.	 Cross-level interaction effects. Does the nature or strength of the relationship between two
lower-level variables (e.g., L1 predictor X and L1 outcome Y) change as a function of a higher-
level variable W? Referring back to our substantive illustration, we are interested in testing the
hypothesis that the relationship between LMX and individual empowerment may vary as a
function of (i.e., is moderated by) the degree of leadership climate such that the relationship will
be stronger for teams with more positive leadership climate and weaker for teams with less
positive leadership climate.




## Datos

```{r}
exdata=read.csv("https://multinivel.netlify.com/practicas/data/aguinis_JOM.csv", header = TRUE, sep = ",")
stargazer(exdata, type = "text")
```

- Rename variables para que sean más coherentes con el ejemplo

```{r}
exdata <- exdata %>% rename(LMX=Xc,Lidclim=Wjc)
```

- Variables relevantes

  - l2id: id de nivel 2, equipos de trabajo
  - LMX: quality of leader–member exchange
  - Lidclim: clima de liderazgo en el equipo
  - Y: variable dependiente, empoderamiento individual

## Modelos

- Modelo Nulo e ICC

```{r}
lmm.fit1=lmer(Y ~ 1 + (1|l2id), data=exdata,REML=F)
summary(lmm.fit1)
reghelper::ICC(lmm.fit1)
```

- Modelo con predictores fijos

```{r}
lmm.fit2=lmer(Y ~1 + LMX + Lidclim +(1|l2id),data=exdata,REML=F)
screenreg(lmm.fit2)
```

- Pendiente aleatoria

```{r}
lmm.fit3=lmer(Y ~1 + LMX + Lidclim +(1 + LMX|l2id), data=exdata,REML=F)
summary(lmm.fit3)
```

Deviance

```{r}
anova(lmm.fit2,lmm.fit3)
```

- Modelo con interacción entre niveles

```{r}
lmm.fit4=lmer(Y ~1 + LMX*Lidclim + (1 + LMX|l2id), data=exdata, REML=F)
screenreg(lmm.fit4)
```


## Reporte final

- Tabla

```{r}
screenreg(list(lmm.fit1, lmm.fit2, lmm.fit3, lmm.fit4))
```


En html

```{r results='asis'}

htmlreg(list(lmm.fit1, lmm.fit2, lmm.fit3, lmm.fit4),
    custom.model.names = c("Nulo","Pendiente <br> fija","Pendiente <br> Aleatoria", "Interacción"),
    custom.coef.names = c("Intercepto", "LMX $(\\gamma_{10})$" ,"Clima liderazgo $(\\gamma_{01})$", "LMX*Clima $(\\gamma_{11})$"),
    custom.gof.names=c(NA,NA,NA,NA,NA,
                   "Var: l2id ($\\tau_{00}$)",
                   "Var: Residual ($\\sigma^2$)",
                   "Var: l2id LMX ($\\tau_{11}$)",
                   "Cov: l2id (Intercept) LMX ($\\tau_{01}$)"),
    custom.note = "%stars. Errores estándar en paréntesis",
    caption="Replicacion Tabla Aguinis",
    caption.above=TRUE,
    digits=3,
    doctype = FALSE)

```


- Interaction Plot


```{r}
plot_model(lmm.fit4, type = "int")
```















# Foro

<script src="https://giscus.app/client.js"
        data-repo="cursos-metodos-facso/multinivel"
        data-repo-id="MDEwOlJlcG9zaXRvcnkyMDA3MTE2MDk="
        data-category="Q&A"
        data-category-id="DIC_kwDOC_aduc4CYhjI"
        data-mapping="title"
        data-strict="0"
        data-reactions-enabled="1"
        data-emit-metadata="0"
        data-input-position="bottom"
        data-theme="light"
        data-lang="es"
        crossorigin="anonymous"
        async>
</script>



